Implementing Equational Constraints\\ in a Functional Language

### 2011-07-13 papers/wlp2011/paper.tex
changed benchmarks to prdfs

 \item compilation into logic languages like Prolog and reusing
the existing backtracking implementation for non-deterministic
search as well as logic variables and unification for computing with partial
information \cite{AntoyHanus00FROCOS,Lopez-FraguasSanchez-Hernandez99}, or

@@

In this paper we consider KiCS2 \cite{BrasselHanusPeemoellerReck11},
a new system that compiles functional logic programs of the
source language Curry \cite{Hanus06Curry}
into purely functional Haskell programs.
We have shown in \cite{BrasselHanusPeemoellerReck11}
that this implementation can compete with or outperform
other existing implementations of Curry.
KiCS2 is based on the idea to represent
the search space, i.e., all non-deterministic results of a computation,
as a data structure that can be traversed by operations
implementing various strategies.
Furthermore, logic variables are replaced by generators,
i.e., operations that non-deterministically evaluate to
all possible ground values of the type of the logic variable.
It has been shown \cite{AntoyHanus06ICLP}
that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
and computing with generators by rewriting
are equivalent, i.e., compute the same values.
Although this implementation technique is correct \cite{Brassel11Thesis},
the generation of all values for logic variables
might be inefficient for applications that
exploit constraints on partial values.
For instance, in Prolog the equality constraint \ccode{X=c(a)}
is solved by instantiating the variable \code{X} to \code{c(a)},
but the equality constraint \ccode{X=Y} is solved by binding \code{X}
to \code{Y} without enumerating any values for \code{X} or \code{Y}.
Therefore, we propose in this paper new techniques
to implement equational constraints in the framework of KiCS2
(note that, in contrast to Prolog, unification is performed
modulo function evaluation).
Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
i.e., patterns containing evaluable operations for more powerful pattern 
matching than in logic or functional languages,
can be implemented in this framework.
We show that both extensions lead to efficiency improvements
without sacrificing the efficiency of the kernel implementation.

In the next section, we review the source language Curry
and the features considered in this paper.
Section~\ref{sec:Compilation} sketches the implementation scheme
of KiCS2.
Sections~\ref{sec:Unification} and~\ref{sec:FuncPatterns}
discuss the extensions to implement unification modulo
functional evaluation and functional patterns, respectively.
Benchmarks demonstrating the usefulness of this scheme
are presented in Section~\ref{sec:Benchmarks}
before we conclude in Section~\ref{sec:Conclusions}.

@@

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{haskell}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  not False        $\to~$ True
\end{haskell}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated prior to the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics},
where values of the arguments of an operation are determined before the
operation is evaluated. This can be enforced in a lazy strategy
by sharing actual arguments.
For instance, the expression above can be lazily evaluated
provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}
consistently.

The condition $c$ in rule (\ref{rule}) typically is a
conjunction of \emph{equational constraints}
of the form \code{$e_1$\,=:=\,$e_2$}.
Such a constraint is satisfiable if
both sides $e_1$ and $e_2$ are reducible to unifiable data terms.
For instance, if the symbol ``\code{++}'' denotes the usual list
concatenation operation, we can define an operation \code{last}
that computes the last element \code{e} of a non-empty list \code{xs}
as follows:
%
\label{ex:last}
\begin{curry}
  last xs | ys++[e] =:= xs  = e   where ys, e free
\end{curry}
%
Like in Haskell, most rules defining functions are \emph{constructor-based} 
\cite{ODonnell85}, i.e., in (\ref{rule})
$t_1,\ldots,t_n$ consist of variables and/or data constructor symbols
only.  However, Curry also allows \emph{functional patterns}
\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might additionally contain calls to
defined operations. For instance, we can also define the last element
of a list by:
%
\begin{curry}
  last' (xs++[e]) = e
\end{curry}
%
Here, the functional pattern \code{(xs++[e])}
states that \code{last'} is reducible
to \code{e} provided that the argument can be matched against
some value of \code{(xs++[e])} where \code{xs} and \code{e} are
free variables.
By instantiating \code{xs} to arbitrary lists, the value
of \code{(xs++[e])} is any list having \code{e} as its last element.
Functional patterns are a powerful feature to express arbitrary 
selections in term structures. For instance, they support a straightforward
processing of XML data with incompletely specified
or evolving formats \cite{Hanus11ICLP}.

@@

As mentioned in the introduction, the KiCS2 implementation
is based on the explicit representation of non-deterministic results
in a data structure.
This is achieved by extending each data type of the source program
by constructors to represent a choice between two values
and a failure, respectively.
For instance, the data type for Boolean values defined in a Curry program by
\begin{curry}
  data Bool = False | True
\end{curry}
is translated into the Haskell data type\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data Bool = False | True | Choice ID Bool Bool | Fail
\end{haskell}
The first argument of type \code{ID} of each \code{Choice} constructor
is used to implement the call-time choice semantics
discussed in Section~\ref{sec:Curry}.
Since the evaluation of \code{xorSelf aBool} duplicates
the argument operation \code{aBool}, we have to ensure
that both duplicates, which later evaluate to a non-deterministic
choice between two values, yield either \code{True} or \code{False}.
This is obtained by assigning a unique identifier (of type \code{ID})
to each \code{Choice}. The difficulty is to get a unique identifier
on demand, i.e., when some operation evaluates to a \code{Choice}.
Since we want to compile into \emph{purely} functional programs
(in order to enable powerful program optimizations),
we cannot use unsafe features with side effects to generate
such identifiers.
Hence, we pass a (conceptually infinite) set of identifiers,
also called \emph{identifier supply},
to each operation so that a \code{Choice} can pick its unique identifier from
this set.
For this purpose, we assume a type \code{IDSupply},
representing an infinite set of identifiers,
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
The operation \code{initSupply} creates such a set (at the beginning
of an execution),
the operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
There are different implementations available \cite{AugustssonRittriSynek94}
(see below for a simple implementation) and our system
is parametric over concrete implementations of \code{IDSupply}.

@@

To show a concrete example, we use the following implementation
of \code{IDSupply} based on unbounded integers:
%
\begin{haskell}
  type IDSupply = Integer
  initSupply    = return 1
  thisID      n = n
  leftSupply  n = 2 * n
  rightSupply n = 2 * n + 1
\end{haskell}
%
If we apply the same transformation to the rules defining \code{xor}
(actually, \code{xor} performs pattern matching on two arguments
which is avoided in KiCS2 by introducing an intermediate operation,
see \cite{BrasselHanusPeemoellerReck11}) and evaluate the main expression
\code{(main 1)},
we obtain the result
\begin{haskell}
  Choice 2 (Choice 2 False True) (Choice 2 True False)
\end{haskell}
Thus, the result is non-deterministic and contains three choices,
whereby all of them have the same identifier.
To extract all values from such a \code{Choice} structure,
we have to traverse it and  compute all possible choices,
but consider the choice identifiers to make consistent (left/right)
decisions.
Thus, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 2 False True)} so that \code{False}
is the only value possible for this branch.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 2 True False)}, which again yields 
\code{False} as the only possible value. In consequence, 
the unintended value \code{True} is not produced.

The requirement to make consistent decisions can be implemented by storing
the decisions already made for some choices during the traversal.
For this purpose, we introduce the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice has
not been decided yet. Furthermore, we assume operations to lookup the 
current decision for a given identifier or change it
(depending on the implementation
of \code{IDSupply}, KiCS2 supports several implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupDecision :: ID -> IO Decision
  setDecision    :: ID -> Decision -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following I/O operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()$\listline$
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
    where
      follow ChooseLeft  = printValsDFS x1
      follow ChooseRight = printValsDFS x2
      follow NoDecision  = do newDecision ChooseLeft  x1
                              newDecision ChooseRight x2$\listline$
      newDecision d x = do setDecision i d
                           printValsDFS x
                           setDecision i NoDecision$\listline$
  printValsDFS v = print v
\end{haskell}
This operation ignores failures and prints values that are not rooted by a
\code{Choice} constructor. For a \code{Choice} constructor, 
it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing all values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

In general, this operation is applied to the normal form
of the main expression (where \code{initSupply} is used to
compute an initial identifier supply passed to this expression).
The normal form computation is necessary for structured data like lists,
so that a failure or choice in some part of the data is moved to
the root.

Other search strategies, like
breadth-first search, iterative deepening, or parallel search,
can be obtained by different implementations of this main operation
to print all values.
Furthermore, one can also collect all values in a tree-like data structure
so that the programmer can implement his own search strategies
(this corresponds to encapsulating search \cite{BrasselHanusHuch04JFLP}).
Finally, instead of printing all values, one can easily define operations
to print either the first solution only or one by one upon user request.
Due to the lazy evaluation strategy of Haskell,
such operations can also be applied to infinite choice structures.

To avoid an unnecessary growth of the search space represented by
\code{Choice} constructors, our compiler performs an optimization for
deterministic operations. If an operation is defined by non-overlapping
rules and does not call, neither directly nor indirectly through
other operations, a function defined by overlapping rules, it does not
introduce non-determinism (like \code{xor} or \code{not}).
Thus, it is not necessary to pass an identifier supply to the operation.
In consequence, only the matching rules are extended, so that the generated code
is nearly identical to a corresponding functional program.
Actually, the benchmarks presented in \cite{BrasselHanusPeemoellerReck11}
show that for deterministic operations this implementation outperforms 
all other Curry implementations,
and, for non-deterministic operations, outperforms Prolog-based
implementations of Curry and can compete with MCC \cite{Lux99FLOPS},
a Curry implementation that compiles to C.

As mentioned in the introduction, occurrences of logic variables are
translated into generators.
For instance, the expression \ccode{not x}, where \code{x} is a logic variable,
is translated into \ccode{not (aBool s)}, where s is an \code{IDSupply}
provided by the context of the expression. The latter expression is 
evaluated by reducing the
argument \code{aBool s} to a choice between \code{True} or \code{False}
followed by applying \code{not} to this choice.
This is similar to a narrowing step on \ccode{not x}
that instantiates the variable \code{x} to \code{True} or \code{False}.
% Actually, it has been shown in \cite{AntoyHanus06ICLP}
% that narrowing expressions with logic variables
% is equivalent to rewriting with generators.
Since such generators are standard
non-deterministic operations, they are translated like any other operation
and therefore do not require any additional run-time support.
However, in the presence of equational constraints,
there are methods which are more efficient than generating all values.
These methods and their implementation are discussed in the
next section.

@@

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & \code{==~}& \code{=:=}& \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   & 134.23    & 0.83      & 0.26        \\
\code{simplify}                         & 8.50      & 7.15      & 7.34        \\
\code{varInExp}                         & 11.72     & 1.50      & 1.40        \\
% \code{varInExp2}                        & 423.01    & 10.50     & 1.04        \\
% \code{varInExp3}                        & 399.68    & 16.70     & 11.28       \\
\code{fromPeano (half (toPeano 10000))} & 32.92     & 6.35      & 11.88       \\
\code{palindrome}                       & 28.59     & 15.58     & 23.21       \\
\code{horseman}                         & 3.58      & 3.13      & n/a         \\
\code{grep}                             & 1.06      & 0.19      & n/a         \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different equation solving methods}
 \label{fig:search-strategies}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & KiCS2 & PAKCS  & MCC  \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.83  & 4.25   & 0.14 \\
\code{simplify}                         & 7.15  & 0.15   & 0.00 \\
\code{varInExp}                         & 1.50  & 3.89   & 0.30 \\
% \code{varInExp2}                        & 10.50 & 3.48   & stckOflw \\
% \code{varInExp3}                        & 16.70 & 19.22  & 1.18 \\
\code{fromPeano (half (toPeano 10000))} & 6.35  & 108.88 & 3.22 \\
\code{palindrome}                       & 15.58 & 32.56  & 1.07 \\
\code{horseman}                         & 3.13  & 8.70   & 0.42 \\
\code{grep}                             & 0.19  & 2.88   & 0.14 \\
\hline
\end{tabular}
\caption{Benchmarks: equational constraints in different Curry implementations}
 \label{fig:unification}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|}
\hline
Expression                              & KiCS2 & PAKCS    \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.26  & 3.97     \\
\code{simplify}                         & 7.34  & 0.27     \\
\code{varInExp}                         & 1.40  & 8.88     \\
% \code{varInExp2}                         & 1.04  & 0.08    \\
% \code{varInExp3}                         & 11.28 & 16.62   \\
\code{fromPeano (half (toPeano 10000))} & 11.88 & $\infty$ \\
\code{palindrome}                       & 23.21 & $\infty$ \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different implementations of functional patterns}
 \label{fig:functional-patterns}
\end{figure}

>>>>

 \item compilation into logic languages (Prolog) and reusing
the existing backtracking implementation for non-deterministic
search as well as logic variables and unification for computing with partial
information \cite{AntoyHanus00FROCOS,Lopez-FraguasSanchez-Hernandez99}, or

@@

In this paper we consider KiCS2 \cite{BrasselHanusPeemoellerReck11},
a new system that compiles functional logic programs of the
source language Curry \cite{Hanus06Curry}
into purely functional Haskell programs.
We have shown in \cite{BrasselHanusPeemoellerReck11}
that this implementation can compete with or outperform
other existing implementations of Curry.
KiCS2 is based on the idea to represent
the search space, i.e., all non-deterministic results of a computation,
as a data structure that can be traversed by operations
implementing various strategies.
Furthermore, logic variables are replaced by generators,
i.e., operations that non-deterministically evaluate to
all possible ground values of the type of the logic variable.
It has been shown \cite{AntoyHanus06ICLP}
that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
or computing with generators by rewriting
are equivalent, i.e., compute the same values.
Although this implementation technique is correct \cite{Brassel11Thesis},
the generation of all values for logic variables
might be inefficient for applications that
exploit constraints on partial values.
For instance, in Prolog the equality constraint \ccode{X=c(a)}
is solved by instantiating the variable \code{X} to \code{c(a)},
but the equality constraint \ccode{X=Y} is solved by binding \code{X}
to \code{Y} without enumerating any values for \code{X} or \code{Y}.
Therefore, we propose in this paper new techniques
to implement equational constraints in the framework of KiCS2
(note that, in contrast to Prolog, unification is performed
modulo function evaluation).
Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
i.e., patterns containing evaluable operations providing
for more powerful pattern matching than in logic or functional languages,
can be implemented in this framework.
We show that both extensions lead to efficiency improvements
without sacrificing the efficiency of the kernel implementation.

In the next section, we review the source language Curry
and the features that are considered in this paper.
Section~\ref{sec:Compilation} sketches the implementation scheme
of KiCS2.
Sections~\ref{sec:Unification} and~\ref{sec:FuncPatterns}
discuss the extensions to implement unification modulo
functional evaluation and functional patterns, respectively.
Benchmarks demonstrating the usefulness of this scheme
are presented in Section~\ref{sec:Benchmarks}
before we conclude in Section~\ref{sec:Conclusions}.

@@

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{haskell}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  not False        $\to~$ True
\end{haskell}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated prior to the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. This can be enforced in a lazy strategy
by sharing actual arguments.
For instance, the expression above can be lazily evaluated
provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}
consistently.

The condition $c$ in rule (\ref{rule}) is typically a
conjunction of \emph{equational constraints}
of the form \code{$e_1$\,=:=\,$e_2$}.
Such a constraint is satisfiable if
both sides $e_1$ and $e_2$ are reducible to unifiable data terms.
For instance, if the symbol ``\code{++}'' denotes the usual list
concatenation operation, we can define an operation \code{last}
that computes the last element \code{e} of a non-empty list \code{xs}
as follows:
%
\label{ex:last}
\begin{curry}
  last xs | ys++[e] =:= xs  = e   where ys, e free
\end{curry}
%
Like in Haskell, most rules defining functions are \emph{constructor-based} 
\cite{ODonnell85}, i.e., in (\ref{rule})
$t_1,\ldots,t_n$ consist of variables and/or data constructor symbols
only.  However, Curry also allows \emph{functional patterns}
\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might additionally contain calls to
defined operations. For instance, we can define the last element
of a list also by:
%
\begin{curry}
  last' (xs++[e]) = e
\end{curry}
%
Here, the functional pattern \code{(xs++[e])}
states that \code{last'} is reducible
to \code{e} provided that the argument can be matched against
some value of \code{(xs++[e])} where \code{xs} and \code{e} are
free variables.
By instantiating \code{xs} to arbitrary lists, the value
of \code{(xs++[e])} is any list having \code{e} as its last element.
Functional patterns are
a powerful feature to express arbitrary selections in term structures.
For instance, functional patterns support a straightforward
processing of XML data with incompletely specified
or evolving formats \cite{Hanus11ICLP}.

@@

As mentioned in the introduction, the KiCS2 implementation
is based on the explicit representation of non-deterministic results
in a data structure.
This is achieved by extending each data type of the source program
by constructors to represent a choice between two values
and a failure, respectively.
For instance, the data type for Boolean values defined in a Curry program by
\begin{curry}
  data Bool = False | True
\end{curry}
is translated into the Haskell data type\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data Bool = False | True | Choice ID Bool Bool | Fail
\end{haskell}
The first argument of type \code{ID} of each \code{Choice} constructor
is used to implement the call-time choice semantics
discussed in Section~\ref{sec:Curry}.
Since the evaluation of \code{xorSelf aBool} duplicates
the argument operation \code{aBool}, we have to ensure
that both duplicates, which later evaluate to a non-deterministic
choice between two values, yield either \code{True} or \code{False}.
This is obtained by assigning a unique identifier (of type \code{ID})
to each \code{Choice}. The difficulty is to get a unique identifier
when needed, i.e., when some operation evaluates to a \code{Choice}.
Since we want to compile into \emph{purely} functional programs
(in order to enable powerful program optimizations),
we cannot use unsafe features with side effects to generate
such identifiers.
Hence, we pass a (conceptually infinite) set of identifiers,
also called \emph{identifier supply},
to each operation so that a \code{Choice} can pick an identifier from
this set.
For this purpose, we assume a type \code{IDSupply},
representing an infinite set of identifiers,
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
The operation \code{initSupply} creates such a set (at the beginning
of an execution),
the operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
There are different implementations available \cite{AugustssonRittriSynek94}
(see below for a simple implementation) and our implementation
is parametric over concrete implementations of \code{IDSupply}.

@@

To show a concrete example, we use the following implementation
of \code{IDSupply} based on unbounded integers:
%
\begin{haskell}
  type IDSupply = Integer
  initSupply    = return 1
  thisID      n = n
  leftSupply  n = 2 * n
  rightSupply n = 2 * n + 1
\end{haskell}
%
If we apply the same transformation to the rules defining \code{xor}
(actually, \code{xor} performs pattern matching on two arguments
which is avoided in KiCS2 by introducing an intermediate operation,
see \cite{BrasselHanusPeemoellerReck11}) and evaluate the main expression
\code{(main 1)},
we obtain the result
\begin{haskell}
  Choice 2 (Choice 2 False True) (Choice 2 True False)
\end{haskell}
Thus, the result is non-deterministic and contains several choices,
but all of them have the same identifier.
This identifier is used when the values contained in this result
are searched for: one has to make \emph{consistent} selections in choices with
same identifiers. Thus, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 2 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 2 True False)} which yields the only
value \code{False}.
Thus, the unintended value \code{True} is not produced.

In general, to extract all values from a \code{Choice} structure,
we have to traverse it, compute all possible choices
but consider the choice identifiers to make consistent (left/right)
decisions.
The latter requirement can be implemented by storing
the decisions already made for some choices during the traversal.
For this purpose, we introduce the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice
is yet undecided.
Furthermore, we assume operations to lookup the current decision
for a given identifier or change its decision (depending on the implementation
of \code{IDSupply}, KiCS2 supports various implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupDecision :: ID -> IO Decision
  setDecision    :: ID -> Decision -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following I/O operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()$\listline$
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
   where
    follow ChooseLeft  = printValsDFS x1
    follow ChooseRight = printValsDFS x2
    follow NoDecision  = do newDecision ChooseLeft  x1
                            newDecision ChooseRight x2$\listline$
    newDecision d x = do setDecision i d
                         printValsDFS x
                         setDecision i NoDecision$\listline$
  printValsDFS v = print v
\end{haskell}
This operation ignores failures and prints values that are not a \code{Choice}.
For a \code{Choice}, it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing
all the values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

In general, this operation is applied to the normal form
of the main expression (where \code{initSupply} is used to
compute an initial identifier supply passed to this expression).
The normal form computation is necessary for structured data like lists
so that a failure or choice in some part of the data is moved to
the root. Other search strategies, like
breadth-first search, iterative deepening, or parallel search,
can be obtained by different implementations of this main operation
to print all values. Instead of printing all values,
one can also collect all values in a tree-like data structure
so that the programmer can implement his own search strategies
(this corresponds to encapsulating search \cite{BrasselHanusHuch04JFLP}).
Finally, one can also have other options rather than printing all values
of the main expression.
For instance, one can easily define operations
to print either the first solution only or one by one upon user request.
Due to the lazy evaluation strategy of Haskell,
such operations can also be applied to infinite choice structures.

To avoid an unnecessary growth of the search space represented by
\code{Choice} constructors, our compiler performs an optimization for
deterministic operations. If an operation does not call, 
neither directly nor indirectly through
other operations, the choice operation \ccode{?},\footnote{%
Note that before applying this compilation scheme, overlapping rules in
user-defined operations are transformed into calls of the choice operation
\ccode{?}.}
then it is not necessary to pass an identifier supply to the operation.
In consequence, only the matching rules are extended so that the generated code
is nearly identical to a corresponding functional program.
Actually, the benchmarks presented in \cite{BrasselHanusPeemoellerReck11}
show that this implementation outperforms all other Curry implementations
for deterministic operations,
and, for non-deterministic operations, outperforms Prolog-based
implementations of Curry and can compete with MCC \cite{Lux99FLOPS},
a Curry implementation that compiles to C.

As mentioned in the introduction,
KiCS2 translates occurrences of logic variables
into generators.
For instance, the expression \ccode{not x}, where \code{x} is a logic variable,
is translated into \ccode{not aBool} (we omit the \code{IDSupply} argument
here). The latter expression is evaluated by reducing the
argument \code{aBool} to a choice between \code{True} or \code{False}
followed by applying \code{not} to this choice.
This is similar to a narrowing step on \ccode{not x}
that instantiates the variable \code{x} to \code{True} or \code{False}.
% Actually, it has been shown in \cite{AntoyHanus06ICLP}
% that narrowing expressions with logic variables
% is equivalent to rewriting with generators.
Since such generators are standard
non-deterministic operations, they are translated like any other operation
so that one does not need additional run-time support for them.
However, if equational constraints should be solved,
there are methods which are more efficient than generating all values.
These methods and their implementation are discussed in the
next section.

@@

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & \code{==~}& \code{=:=}& \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   &  2.91   &    0.05   &    0.01   \\
\code{simplify}                         & 10.30      & 6.77      & 7.07        \\
\code{varInExp}                         & 2.34    & 0.24      & 0.21        \\
% \code{varInExp2}                        & 0.34   & 0.02     & 0.01        \\
% \code{varInExp3}                        & 13.28    & 1.36     & 1.02       \\
\code{fromPeano (half (toPeano 10000))} & 26.67     & 5.95      & 11.19       \\
\code{palindrome}                       & 30.86     & 14.05     & 20.26       \\
\code{horseman}                         & 3.24      & 3.31      & n/a         \\
\code{grep}                             & 1.06      & 0.10      & n/a         \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different equation solving methods}
 \label{fig:search-strategies}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & KiCS2 & PAKCS  & MCC  \\
\hline
\code{last (map (inc 0) [1..1000])}   & 0.05  & 0.40   &  0.01\\
\code{simplify}                         & 6.77  & 0.15   & 0.00 \\
\code{varInExp}                         & 0.24  & 0.89   & 0.07 \\
% \code{varInExp2}                        & 0.02 & 0.06   & 0.00 \\
% \code{varInExp3}                        & 1.36 & 2.07  & 0.14 \\
\code{fromPeano (half (toPeano 10000))} & 5.95  & 108.88 & 3.22 \\
\code{palindrome}                       & 14.05 & 32.56  & 1.07 \\
\code{horseman}                         & 3.31  & 8.70   & 0.42 \\
\code{grep}                             & 0.10  & 2.88   & 0.14 \\
\hline
\end{tabular}
\caption{Benchmarks: equational constraints in different Curry implementations}
 \label{fig:unification}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|}
\hline
Expression                              & KiCS2 & PAKCS    \\
\hline
\code{last (map (inc 0) [1..10000])}   &  0.01 & 0.33     \\
\code{simplify}                         & 7.07  & 0.27     \\
\code{varInExp}                         &  0.21  &  1.87    \\
% \code{varInExp2}                         & 0.01  & 0.02    \\
% \code{varInExp3}                         & 1.02 & 2.16   \\
\code{fromPeano (half (toPeano 10000))} & 11.19 & $\infty$ \\
\code{palindrome}                       & 20.26 & $\infty$ \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different implementations of functional patterns}
 \label{fig:functional-patterns}
\end{figure}
